---
title: "project-final"
author: "Gabriele Chignoli"
date: "12/21/2017"
output: html_document
---

## Do formants depend on speakers ?
***

### 1st step

Values of vowel formants can be a great indicator in order to distinguish between different vocalic sounds, but when it comes to voice recognition can formants help us to *recognize* different speakers and if they do, which are the more relevant ? We will analyse the first three formants, their distributions and their variations in order to try to answer this preliminary question. *F1* and *f2* values mainly corresponding to the average values of *ideal vowels*, in Automatic Speaker Recognition systems the *f3* has in particular a significant role to identify, for exemple in comparaison pairs, if it is the same person who's speaking or two different ones.  
The other great value we will take in account is the pitch, noted *f0*. In human language the pitch can play a heavy role in recogniziting the speaker native language, the mood of a person and in some cases it could be the element which to characterise a certain speaker.
So where do we start ? By loading the data and making some assumption on how we can manipulate them.

```{r loadDataset}
voyelles = read.csv("data.csv", header = TRUE, sep = "\t")
```

### Dataset

Our dataset is made of 27 records by 9 different speakers, 5 women and 4 men, french native speakers. Every participant read 55 words, reapeting the set 3 times. All the values have been extracted automatically with Praat and saved in the table from the previous chunk, we have 6 columns in total : speaker gender, vowel, pitch value, first formants, second formant and third formant.

```{r vocalicTrianglePlot}
ggplot2::ggplot(voyelles, ggplot2::aes(F2, F1)) +
  ggplot2::geom_point(ggplot2::aes(col=VOWEL), size=3) +
  ggplot2::scale_colour_brewer(palette = "Spectral") +
  ggplot2::scale_x_reverse() + ggplot2::scale_y_reverse() +
  ggplot2::labs(y="F1", x="F2", title="Dataset vocalic triangle")
```

Not all french vowels are produced in the dataset but it will not have a great impact on our question, the vowels we consider are /a/ /u/ /i/ and we find in "@" all kind of schwa or *coup de glotte* (this explains the large distribution it has on the triangle).  
In many cases /u/ and /i/ are closer to the values of /y/ but the original annotation doesn't contains any /y/.

### How to answer to the question

##### 1. F3 values

As we said in the introduction, the first element which can lead us to recognize if it is a speaker or multiple one to be analysed in a dataset would be the third formant values of that speaker. Here it is a global *f3* representation.   
At first look we can't tell if a great variation is present in our data, there are in every vowels some areas with a greater concentration of values but we don't have an explicit explaination to what this could be. We can start making two hypothesis :

- all the areas of concentration are common to avery speaker so that for a single speaker we will have a plot pretty much similar to this one. In this case we would assume that the *f3* can't be a reliable indicator of variation ;
- the plot we see here is just the addition of all the *f3* belonging to the different speakers. This is the hypothesis we'll try to demonstrate in order to confirm the statements made in the introduction.

```{r f3Plot}
ggplot2::ggplot(voyelles, ggplot2::aes(VOWEL, F3)) +
  ggplot2::geom_count(col="tomato3", show.legend=F) +
  ggplot2::labs(y="F3", title="Dataset F3 values") 
```

##### 2. F0 values

The second indicator we will consider in order to identify different speakers will be the pitch or f0 measure. We'll present different representations, all based on density. We won't look at the values of this measure, they will not tell us something in particular, but just looking at the shapes of density curves we will understand the great variation of our dataset.

```{r pithPlotGenders}
ggplot2::ggplot(voyelles, ggplot2::aes(F0)) +
  ggplot2::geom_density(ggplot2::aes(fill=as.factor(SPEAKER)), alpha=0.8) +
  ggplot2::scale_colour_brewer(palette = "Spectral") +
  ggplot2::labs(title="Pitch measurements", fill="SPEAKER")
```

+ The first one is based only only on gender variation, the two curves tell us one simple thing : the speaker gender has a great impact on the pitch measure (not a great discover). Men's pitch is lower than women's one, due to the physiological difference between men and women frequencies, cavities, this tells us that just in thanks to pitch measurement we would, in general, predict if it's a male voice or a female one.

```{r pitchPlotVowels}
ggplot2::ggplot(voyelles, ggplot2::aes(F0)) +
  ggplot2::geom_density(ggplot2::aes(fill=as.factor(VOWEL)), alpha=0.8) +
  ggplot2::scale_colour_brewer(palette = "Spectral") +
  ggplot2::labs(title="Pitch measurements", fill="VOWEL") + ggplot2::facet_wrap( ~ VOWEL)
```

+ We applied the density measurement taking the vowels as a variation factor. this time our curves differs from one another but we can't really see some important information from them. Considering the global realisation of a vowel knowing that the pitch is strongly influenced by gender lead us to a not really satisfying representation of values.

```{r pitchPlotVowelsGenders}
ggplot2::ggplot(voyelles, ggplot2::aes(F0)) +
  ggplot2::geom_density(ggplot2::aes(fill=as.factor(SPEAKER)), alpha=0.8) +
  ggplot2::scale_colour_brewer(palette = "Spectral") +
  ggplot2::labs(title="Pitch measurements", fill="VOWEL") + ggplot2::facet_wrap( ~ VOWEL)
```

+ In order to clarify the variation between vowel we add gender as a factor of variation in this new density representation. As we see, this time our curves directly derive from our first pitch curve, but as that one it can help us to describe the variation between genders not specifically between speakers.  

##### 3. Cavity lengths  

Another factor we could use to predict the speaker variation can be the cavity length ? Maybe this measure can be and addiction to other more precise ones :  
* because of physiological factors, the human cavity is almost the same for every being, this measure isn't really specific for a person ;  
* it isn't always precide, it can vary because we just computed it from formants measurements, using the reverse formant calcutation formula $Fn = (2n-1)c/4l$ where Fn is the sum of all formants or a specific formant l is the length and c the constant of sound speed (35000 m/s) so the length formula will be  $l = (c/Fn)/4$
+ We applied the density measurement taking the vowels as a variation factor. this time our curves differs from one another but we can't really see some important information from them. Considering the global realisation of a vowel knowing that the pitch is strongly influenced by gender lead us to a not really satisfying representation of values.

```{r cavityLength, echo=FALSE}
cavityF1 <- c()
cavityF2 <- c()
cavityF3 <- c()
cavities <- c()

for (i in voyelles[["F1"]] ) {
    l <- (35000/i)/4
    cavityF1 <- append(cavityF1, l)
}
for (i in voyelles[["F2"]] ) {
    l <- (35000/i)/4
    cavityF2 <- append(cavityF2, l)
}
for (i in voyelles[["F3"]] ) {
    l <- (35000/i)/4
    cavityF3 <- append(cavityF3, l)
}
for (i in 1:1885) {
  lm <- (cavityF1[i] + cavityF2[i] + cavityF3[i])/3
  cavities <-(append(cavities, lm))
}
```

```{r cavityLengthPlot}
ggplot2::ggplot(data.frame(cavities), ggplot2::aes(seq_along(cavities),cavities)) + 
  ggplot2::geom_bar(colour="black", stat="identity") + 
  ggplot2::labs(y="length", x="count", title="Average cavity length")
```

***
#### Single Speaker considerations

In order to realize our main idea to recognize between two or more different speakers we will now compare one pair of speakers for every gender, using the *f3* and *f0* measures.

```{r subset}
voyellesLocutrice1 <- voyelles[1:168,]
voyellesLocutrice2 <- voyelles[169:337,]
voyellesLocuteur1 <- voyelles[1497:1720,]
voyellesLocuteur2 <- voyelles[1719:1885,]
```

##### 1. Vocalic triangle

```{r vocalicTriangleCompW}
ggplot2::ggplot(voyellesLocutrice1, ggplot2::aes(x=F2, y=F1)) +
  ggplot2::geom_point(ggplot2::aes(col=VOWEL), size=3) +
  ggplot2::scale_colour_brewer(palette = "Spectral") +
  ggplot2::scale_x_reverse() + ggplot2::scale_y_reverse() +
  ggplot2::labs(y="F1", x="F2", title="Female speaker 1 vocalic triangle")

ggplot2::ggplot(voyellesLocutrice2, ggplot2::aes(x=F2, y=F1)) +
  ggplot2::geom_point(ggplot2::aes(col=VOWEL), size=3) +
  ggplot2::scale_colour_brewer(palette = "Spectral") +
  ggplot2::scale_x_reverse() + ggplot2::scale_y_reverse() +
  ggplot2::labs(y="F1", x="F2", title="Female speaker 2 vocalic triangle")
```

##### 2. Pitch measurement

```{r pitchCompW}
ggplot2::ggplot(voyellesLocutrice1, ggplot2::aes(F0)) +
  ggplot2::geom_density(ggplot2::aes(fill=as.factor(SPEAKER)), alpha=0.8) +
  ggplot2::scale_colour_brewer(palette = "Spectral") +
  ggplot2::labs(title="Pitch measurements female speaker 1", fill="GENDER")

ggplot2::ggplot(voyellesLocutrice2, ggplot2::aes(F0)) +
  ggplot2::geom_density(ggplot2::aes(fill=as.factor(SPEAKER)), alpha=0.8) +
  ggplot2::scale_colour_brewer(palette = "Spectral") +
  ggplot2::labs(title="Pitch measurements female speaker 2", fill="GENDER")
```

##### 3. F3 comparaison

```{r f3CompW}
ggplot2::ggplot(voyellesLocutrice1, ggplot2::aes(VOWEL, F3)) +
  ggplot2::geom_count(col="tomato3", show.legend=F) +
  ggplot2::labs(y="F3", title="Female speaker 1 F3 values")

ggplot2::ggplot(voyellesLocutrice2, ggplot2::aes(VOWEL, F3)) +
  ggplot2::geom_count(col="tomato3", show.legend=F) +
  ggplot2::labs(y="F3", title="Female speaker 2 F3 values")
```

##### 1. Vocalic triangle

```{r vocalicTriangleCompM}
ggplot2::ggplot(voyellesLocuteur1, ggplot2::aes(x=F2, y=F1)) +
  ggplot2::geom_point(ggplot2::aes(col=VOWEL), size=3) +
  ggplot2::scale_colour_brewer(palette = "Spectral") +
  ggplot2::scale_x_reverse() + ggplot2::scale_y_reverse() +
  ggplot2::labs(y="F1", x="F2", title="Male speaker 1 vocalic triangle")

ggplot2::ggplot(voyellesLocuteur2, ggplot2::aes(x=F2, y=F1)) +
  ggplot2::geom_point(ggplot2::aes(col=VOWEL), size=3) +
  ggplot2::scale_colour_brewer(palette = "Spectral") +
  ggplot2::scale_x_reverse() + ggplot2::scale_y_reverse() +
  ggplot2::labs(y="F1", x="F2", title="Male speaker 2 vocalic triangle")
```

##### 2. Pitch measurement

```{r pitchCompM}
ggplot2::ggplot(voyellesLocuteur1, ggplot2::aes(F0)) +
  ggplot2::geom_density(ggplot2::aes(fill=as.factor(SPEAKER)), alpha=0.8) +
  ggplot2::scale_colour_brewer(palette = "Spectral") +
  ggplot2::labs(title="Pitch measurements Male speaker 1", fill="GENDER")

ggplot2::ggplot(voyellesLocuteur2, ggplot2::aes(F0)) +
  ggplot2::geom_density(ggplot2::aes(fill=as.factor(SPEAKER)), alpha=0.8) +
  ggplot2::scale_colour_brewer(palette = "Spectral") +
  ggplot2::labs(title="Pitch measurements Male speaker 2", fill="GENDER")
```

##### 3. F3 comparaison

```{r f3CompM}
ggplot2::ggplot(voyellesLocuteur1, ggplot2::aes(VOWEL, F3)) +
  ggplot2::geom_count(col="tomato3", show.legend=F) +
  ggplot2::labs(y="F3", title="Male speaker 1 F3 values")

ggplot2::ggplot(voyellesLocuteur2, ggplot2::aes(VOWEL, F3)) +
  ggplot2::geom_count(col="tomato3", show.legend=F) +
  ggplot2::labs(y="F3", title="Male speaker 2 F3 values")
```
