---
title: "project-final"
author: "Gabriele Chignoli"
date: "12/21/2017"
output: html_document
---

## Do formants depend on speakers ?

### 1st step

Values of vowel formants can be a great indicator in order to distinguish between different vocalic sounds, but when it comes to voice recognition can formants help us to *recognize* different speakers and if they do, which are the more relevant ? We will analyse the first three formants, their distributions and their variations in order to try to answer this preliminary question. *F1* and *f2* values mainly corresponding to the average values of *ideal vowels*, in Automatic Speaker Recognition systems the *f3* has in particular a significant role to identify, for exemple in comparaison pairs, if it is the same person who's speaking or two different ones.  
The other great value we will take in account is the pitch, noted *f0*. In human language the pitch can play a heavy role in recogniziting the speaker native language, the mood of a person and in some cases it could be the element which to characterise a certain speaker.
So where do we start ? By loading the data and making some assumption on how we can manipulate them.

```{r loadDataset}
voyelles = read.csv("data.csv", header = TRUE, sep = "\t")
```

### Dataset

Our dataset is made of 27 records by 9 different speakers, 5 women and 4 men, french native speakers. Every participant read 55 words, reapeting the set 3 times. All the values have been extracted automatically with Praat and saved in the table from the previous chunk, we have 6 columns in total : speaker gender, vowel, pitch value, first formants, second formant and third formant.

```{r vocalicTrianglePlot}
ggplot2::ggplot(voyelles, ggplot2::aes(F2, F1)) +
  ggplot2::geom_point(ggplot2::aes(col=VOWEL), size=3) +
  ggplot2::scale_colour_brewer(palette = "Spectral") +
  ggplot2::scale_x_reverse() + ggplot2::scale_y_reverse() +
  ggplot2::labs(y="F1", x="F2", title="Dataset vocalic triangle")
```

Not all french vowels are produced in the dataset but it will not have a great impact on our question, the vowels we consider are /a/ /u/ /i/ and we find in "@" all kind of schwa or *coup de glotte* (this explains the large distribution it has on the triangle).  
In many cases /u/ and /i/ are closer to the values of /y/ but the original annotation doesn't contains any /y/.

### How to answer to the question

##### 1. F3 values

As we said in the introduction, the first element which can lead us to recognize if it is a speaker or multiple one to be analysed in a dataset would be the third formant values of that speaker. Here it is a global *f3* representation.   
At first look we can't tell if a great variation is present in our data, there are in every vowels some areas with a greater concentration of values but we don't have an explicit explaination to what this could be. We can start making two hypothesis :

- all the areas of concentration are common to avery speaker so that for a single speaker we will have a plot pretty much similar to this one. In this case we would assume that the *f3* can't be a reliable indicator of variation ;
- the plot we see here is just the addition of all the *f3* belonging to the different speakers. This is the hypothesis we'll try to demonstrate in order to confirm the statements made in the introduction.

```{r f3Plot}
ggplot2::ggplot(voyelles, ggplot2::aes(VOWEL, F3)) +
  ggplot2::geom_count(col="tomato3", show.legend=F) +
  ggplot2::labs(y="F3", title="Dataset F3 values") 
```

##### 2. F0 values

The second indicator we will consider in order to identify different speakers will be the pitch or f0 measure.

```{r pitchPlot}
ggplot2::ggplot(voyelles, ggplot2::aes(F0)) +
  ggplot2::geom_density(ggplot2::aes(fill=as.factor(VOWEL)), alpha=0.8) +
  ggplot2::scale_colour_brewer(palette = "Spectral") +
  ggplot2::labs(title="Pitch measurements", fill="VOWEL") + ggplot2::facet_wrap( ~ VOWEL)
```

```{r}
ggplot2::ggplot(voyelles, ggplot2::aes(F0)) +
  ggplot2::geom_density(ggplot2::aes(fill=as.factor(SPEAKER)), alpha=0.8) +
  ggplot2::scale_colour_brewer(palette = "Spectral") +
  ggplot2::labs(title="Pitch measurements", fill="SPEAKER")
```

##### 3. Cavity lengths

```{r cavityLength}
cavityF1 <- c()
cavityF2 <- c()
cavityF3 <- c()
cavities <- c()

for (i in voyelles[["F1"]] ) {
    l <- (35000/i)/4
    cavityF1 <- append(cavityF1, l)
}
for (i in voyelles[["F2"]] ) {
    l <- (35000/i)/4
    cavityF2 <- append(cavityF2, l)
}
for (i in voyelles[["F3"]] ) {
    l <- (35000/i)/4
    cavityF3 <- append(cavityF3, l)
}
for (i in 1:1885) {
  lm <- (cavityF1[i] + cavityF2[i] + cavityF3[i])/3
  cavities <-(append(cavities, lm))
}

ggplot2::ggplot(data.frame(cavities), ggplot2::aes(seq_along(cavities),cavities)) + 
  ggplot2::geom_bar(colour="black", stat="identity") + 
  ggplot2::labs(y="length", x="count", title="Average cavity length")
```


#### Single Speaker considerations

```{r subsetOnly1SpeakerW1}
voyelles1Locutrice1 <- voyelles[1:168,]
ggplot2::ggplot(voyelles1Locutrice1, ggplot2::aes(x=F2, y=F1)) +
  ggplot2::geom_point(ggplot2::aes(col=VOWEL), size=3) +
  ggplot2::scale_colour_brewer(palette = "Spectral") +
  ggplot2::scale_x_reverse() + ggplot2::scale_y_reverse() +
  ggplot2::labs(y="F1", x="F2", title="1 female speaker vocalic triangle")


ggplot2::ggplot(voyelles1Locutrice1, ggplot2::aes(F0)) +
  ggplot2::geom_density(ggplot2::aes(fill=as.factor(SPEAKER)), alpha=0.8) +
  ggplot2::scale_colour_brewer(palette = "Spectral") +
  ggplot2::labs(title="Pitch measurements single female speaker", fill="GENDER")

ggplot2::ggplot(voyelles1Locutrice1, ggplot2::aes(VOWEL, F3)) +
  ggplot2::geom_count(col="tomato3", show.legend=F) +
  ggplot2::labs(y="F3", title="Single female speaker F3 values")
```

```{r subsetOnly1SpeakerW2}
voyelles1Locutrice2 <- voyelles[169:337,]
ggplot2::ggplot(voyelles1Locutrice2, ggplot2::aes(x=F2, y=F1)) +
  ggplot2::geom_point(ggplot2::aes(col=VOWEL), size=3) +
  ggplot2::scale_colour_brewer(palette = "Spectral") +
  ggplot2::scale_x_reverse() + ggplot2::scale_y_reverse() +
  ggplot2::labs(y="F1", x="F2", title="1 female speaker vocalic triangle")


ggplot2::ggplot(voyelles1Locutrice2, ggplot2::aes(F0)) +
  ggplot2::geom_density(ggplot2::aes(fill=as.factor(SPEAKER)), alpha=0.8) +
  ggplot2::scale_colour_brewer(palette = "Spectral") +
  ggplot2::labs(title="Pitch measurements single female speaker", fill="GENDER")

ggplot2::ggplot(voyelles1Locutrice2, ggplot2::aes(VOWEL, F3)) +
  ggplot2::geom_count(col="tomato3", show.legend=F) +
  ggplot2::labs(y="F3", title="Single female speaker F3 values")
```

```{r subsetOnly1SpeakerM1}
voyelles1Locuteur1 <- voyelles[1497:1720,]
ggplot2::ggplot(voyelles1Locuteur1, ggplot2::aes(x=F2, y=F1)) +
  ggplot2::geom_point(ggplot2::aes(col=VOWEL), size=3) +
  ggplot2::scale_colour_brewer(palette = "Spectral") +
  ggplot2::scale_x_reverse() + ggplot2::scale_y_reverse() +
  ggplot2::labs(y="F1", x="F2", title="1 male speaker vocalic triangle")

ggplot2::ggplot(voyelles1Locuteur1, ggplot2::aes(F0)) +
  ggplot2::geom_density(ggplot2::aes(fill=as.factor(SPEAKER)), alpha=0.8) +
  ggplot2::scale_colour_brewer(palette = "Spectral") +
  ggplot2::labs(title="Pitch measurements single male speaker", fill="GENDER")

ggplot2::ggplot(voyelles1Locuteur1, ggplot2::aes(VOWEL, F3)) +
  ggplot2::geom_count(col="tomato3", show.legend=F) +
  ggplot2::labs(y="F3", title="Single male speaker F3 values")
```

```{r subsetOnly1SpeakerM2}
voyelles1Locuteur2 <- voyelles[1719:1885,]
ggplot2::ggplot(voyelles1Locuteur2, ggplot2::aes(x=F2, y=F1)) +
  ggplot2::geom_point(ggplot2::aes(col=VOWEL), size=3) +
  ggplot2::scale_colour_brewer(palette = "Spectral") +
  ggplot2::scale_x_reverse() + ggplot2::scale_y_reverse() +
  ggplot2::labs(y="F1", x="F2", title="1 male speaker vocalic triangle")

ggplot2::ggplot(voyelles1Locuteur2, ggplot2::aes(F0)) +
  ggplot2::geom_density(ggplot2::aes(fill=as.factor(SPEAKER)), alpha=0.8) +
  ggplot2::scale_colour_brewer(palette = "Spectral") +
  ggplot2::labs(title="Pitch measurements single male speaker", fill="GENDER")

ggplot2::ggplot(voyelles1Locuteur2, ggplot2::aes(VOWEL, F3)) +
  ggplot2::geom_count(col="tomato3", show.legend=F) +
  ggplot2::labs(y="F3", title="Single male speaker F3 values")
```

